A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-02-26 paper/bare_conf.tex
Add stuff to the PyOP2 compiler section

Optimization of assembly kernels is non-trivial. Polyhedral compilation could be used to improve performance; however, given the exceptionally small size of the kernels, effective register locality can be obtained only by an in-depth analysis of the data space, rather than reasoning on the iteration space (our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, were unsuccessful). BLAS routines could be employed as well, although fairly complicated code manipulations would be required, for a performance gain potentially far from the expectations. The major issue concerns automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of the memory access pattern. Register locality would also be affected, since the same matrix could appear in a number of products. BLAS libraries are also known to perform distant from peak performance when the dimension of the matrices is small~\cite{nek5000}. Hand-made BLAS-based implementations of the Helmholtz assembly kernel (illustrated later), varying the polynomial order from $p=1$ to $p=4$, has showed that the performance is worse than that achieved with our optimization strategy.

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. This high-level specification is translated by the domain-specific Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel. ASTs are then passed to PyOP2~\cite{pyop2isc}, which lies at the core of Firedrake, where parallel execution over the unstructured mesh is managed. Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it receives FFC's ASTs as input, it introduces optimizations, and it generates C code as output, which is eventually just-in-time compiled on the underlying architecture.

The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}. Initially, an AST is inspected, looking for the presence of iteration spaces and other domain-specific information provided by the higher layer. If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied. This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{reguly-inter-kernel}. The second transformation step is applied if the backend is a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{quadrature-cuda}. Then, a sequence of optimization steps are executed. A cost model is queried to estimate the potential gain if the $split$ and $op$-$vect$ transformations are applied. Optimization passes must follow a specific order. Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays. Then, $split$ might take place, depending on the outcome of the cost model. The $op$-$vect$ optimization is introduced lastly. Note that the compiler can handle all possible corner cases: for example, if $op$-$vect$ is to be applied but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.

Our experience... unrolling/unroll and jam/interchange

>>>>

Optimization of assembly kernels is non-trivial. Polyhedral compilation could be used to improve performance; however, given the exceptionally small size of the kernels, effective register locality can be obtained only through an in-depth analysis of the data space, rather than by reasoning on the iteration space (our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, did not improve performance over $icc$ compilation). BLAS routines could be employed as well, although fairly complicated code manipulation would be required, for a performance gain potentially far from the expectation. The major issue would concern automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of control- and data-flow. BLAS libraries are also known to perform distant from peak performance when the dimension of the matrices is small~\cite{nek5000}. In addition, register locality would be affected, since the same matrix could appear in a number of products. Hand-made BLAS-based implementations of the Helmholtz assembly kernel (illustrated later), varying the polynomial order from $p=1$ to $p=4$, has showed that the performance is worse than that achieved with our optimization strategy.

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. This high-level specification is translated by the Fenics Form Compiler~\cite{ffc} into an Abstract Syntax Tree representation of a Finite Element assembly kernel. ASTs are then passed to PyOP2~\cite{pyop2isc}, which lies at the core of Firedrake, where parallel execution over the unstructured mesh is managed. Our compiler, capable of applying the transformations described in Section~\ref{sec:code-transf}, is integrated with PyOP2: it receives FFC's ASTs as input, it introduces optimizations, and it generates C code as output, which is eventually just-in-time compiled on the underlying architecture.

TODO: say there is a very huge space of implementation opportunities; in this section, we say how to prune it (without autotuning, which would be prohibitive for our jit-compilation).

The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}. Initially, an AST is inspected, looking for the presence of iteration spaces and other domain-specific information provided by the higher layer. If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied. This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{reguly-inter-kernel}. The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{quadrature-cuda}. Then, a sequence of optimization steps are executed. A cost model is queried to estimate the potential gain if the $split$ and $op$-$vect$ transformations were applied. Optimization passes must follow a specific order. Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays. Then, $split$ might take place, depending on the outcome of the cost model. The $op$-$vect$ optimization is introduced lastly. Note that the compiler can handle all possible corner cases: for example, if $op$-$vect$ is to be applied but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.

Loop unroll and unroll-and-jam of outer loops are fundamental to expose ILP and data reuse, and so tuning critical parameters such as the unroll factor becomes of great importance. Autotuning might be used, although we avoid it to minimize the overhead of Firedrake's just-in-time compilation. It is our experience (inspection of assembly code, comparison with other hand-made implementations), however, that for our assembly kernels, where the loop nest is affine, bounds are known at compile-time, and memory accesses are unit-stride, newest versions of a vendor compiler like $icc$ employ cost models capable of estimating close-to-optimal values for such parameters.  We therefore leave the backend compiler in charge to select unroll and unroll-and-jam factors. This choice also simplifies the PyOP2 compiler's cost model. The only situation in which we explicitly unroll-and-jam a loop is when $op$-$vect$ is used, since the transformed code seems to prevent the $icc$ compiler from applying any other unroll-based optimizations.

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. It would be possible, therefore, to permute loops in order to improve data movement, although we discovered that a specific loop order often leads to the best run-time, for the storage layout we employ. The loop order $ijk$ is likely to maximize performance. Conceptually, this is motivated by the following considerations: 1) all of the accessed arrays depend on the $i$ loop, and having it in one of the inner positions would mean a significantly higher number of load operations, with a performance loss greater than the reduction of store operations into the stiffness matrix; 2) if loop-invariant code motion is applied, then the number of loads along the $j$ and $k$ dimensions becomes identical (this property derives from the nature of assembly), which implies the loop order $ijk$ is performance-wise identical to $ikj$. These observations, which allow us to fix a specific loop order, contribute to relieve the decisions the PyOP2 compiler has to make.
